TTS2016R: A data set to study population and employment patterns from the 2016 Transportation Tomorrow Survey (TTS) in the Greater Toronto and Hamilton Area in Canada

### 2022-03-11 Manuscript-Data-Package/Manuscript-Data-Package.Rmd
Final edits to manuscript and a gramatical fix to package contents.

---
title: "TTS2016R: A data set to study population and employment patterns from the 2016 Transportation Tomorrow Survey (TTS) in the Greater Toronto and Hamilton Area in Canada"
author:
  - name: "Anastasia Soukhov"
    email: "soukhoa@mcmaster.ca"
    affiliation: SEES
  - name: "Antonio Paez"
    email: "paezha@mcmaster.ca"
    affiliation: SEES
    footnote: 1
address:
  - code: SEES
    address: "School of Earth, Environment and Society, McMaster University, Hamilton, ON, L8S 4K1, Canada"
footnote:
  - code: 1
    text: "Corresponding Author"
abstract: 
  "This paper describes and visualises the data contained within the {TTS2016R} data-package created in `R`, the statistical computing and graphics language. In addition to a synthetic example, {TTS2016R} contains home-to-work commute information for the Greater Golden Horseshoe area in Canada retrieved from the 2016 Transportation Tomorrow Survey (TTS). Included are all Traffic Analysis Zones (TAZ), the number of people who are employed full-time per TAZ, the number of jobs per TAZ, and the origin and destination trips. {TTS2016R} also includes the calculated car travel time from TAZ origin-destination centroid pairs. This information is useful to understand patterns of commuting in the region. We illustrate the estimation of a distance-decay curve using these data. {TTS2016R} can be freely downloaded and explored at: https://github.com/soukhova/TTS2016R where the documentation and code involved in data creation, manupulation, and the final data products are detailed."

@@

```{r load-packages, include=FALSE, cache=FALSE}
library(TTS2016R)
library(dplyr)
library(fitdistrplus)
library(ggplot2)
library(kableExtra)
library(patchwork)
library(sf)
library(scales)
library(stats)
library(ggspatial)
library(shadowtext)
library(tmap)
# library(ggpmisc)
# library(ggrepel)
# library(cowplot)
# library(spdep)
# library(RColorBrewer)
# library(extrafont)
# font_import()
# loadfonts(device = "win")

@@

This manuscript presents the open data product [{TTS2016R}](https://github.com/soukhova/TTS2016R). Open data products are the result of turning source data (open or otherwise) into accessible information that adds value to the original inputs [see @Arribas2021open]. The product presented in this paper is an `R` data package which currently consists of three objects which are sourced from the 2016 Transportation Tomorrow Survey (TTS) or are curated to facilitate the use and analysis of TTS data. This package includes person-to-jobs origin-destinations, traffic analysis zone (TAZ) boundaries, and planning/municipality boundaries for the Greater Golden Horse area (GGH) located in southern Ontario, Canada [@data_management_group_tts_2018]. In addition, the package includes TAZ centroid-to-centroid travel times by car computed using package {r5r} [@Pereira2021r5r]. The aim of this paper is to walk readers through the empirical home-based work commute data set, illustrate the calculation of an impedance function, showcase how it can be used to calculate accessibility to employment, and invite its use in other applications.

Data from the TTS are in principle available to the public but are not fully open, since permission to access the data retrieval system is required. In addition, the raw data can be technically demanding, cumbersome to work with, and requires multiple software to process. By pre-processing the data in a `R` environment, {TTS2016R} offers a slice of the TTS data useful to understand patterns of commuting to work in the region. It also provides open infrastructure for additional TTS or complimentary data sets to be amended by the authors or wider open-source community in the future.

@@

{TTS2016R} includes counts of fully-employed population by place of residence and counts of full-time jobs by place of work aggregated by TAZ (n=`r round(length(TTS2016R::ggh_taz$GTA06), 3) %>% prettyNum(big.mark = ",")` within the survey boundaries). Unique identifiers use the GTA06 Zoning System of TTS. The number of jobs (`r round(sum(TTS2016R::ggh_taz$jobs), 3) %>% prettyNum(big.mark = ",")`) and workers (`r round(sum(TTS2016R::ggh_taz$workers), 3) %>% prettyNum(big.mark = ",")`) are organized in the form of an origin-destination table, indicative of patterns of home-to-work commute (there are `r round(sum(TTS2016R::od_ft_tt$trips), 3) %>% prettyNum(big.mark = ",")` potential interactions). These data were retrieved from the Transportation Tomorrow Survey Data Retrieval System on October 28, 2021 and reflect the potential interaction of full-time employed people and jobs within the GGH survey boundaries shown in Figure \ref{fig:TTS-16-survey-area} as defined by the 2016 TTS methodology [@data_management_group_tts_2018].

Also included in {TTS2016R} are travel times and cost of travel from origin to destination by car; travel times are calculated using the `R` package {r5r}. These travel times are useful to estimate the cost of travel and to calculate impedance functions, among other possible uses. It is important to note that for simplicity, all interactions within {TTS2016R} are assumed to be taken by car, and the travel time is calculated from an origin TAZ centroid to a destination TAZ centroid. The centroid is snapped to the nearest street line by `r5r` and the travel time is calculated for all trips assuming a car travel mode. Additionally, only travel times less than or equal to 180 mins (3 hrs) are calculated; this threshold represents 99% of trip's travel times which are summarized in the descriptive statistics in Table \ref{tab:TTS-16-desc-stats}.

@@

group_ggh_pd[group_ggh_pd$REGION_name=="Peterborough County", "Y"] <- 4921000
```
```{r, echo=FALSE, message=FALSE, warning=FALSE, include = FALSE, out.height=400}
#plot 
ggplot() +
  geom_sf(data = TTS2016R::ggh_pd,
          fill = "gainsboro", color = "darkgray") +
  geom_sf(data = group_ggh_pd_poly, 
          color = "steelblue", fill = NA, size = 0.7) +
  annotation_north_arrow(location = "tl", # north arrow for both the main plot and inset
                         height = unit(0.8, "cm"), 
                         width = unit(0.8, "cm"),
                         style = north_arrow_orienteering(line_width = 0.25,
                                                          line_col = "dimgrey", 
                                                          fill = c("white","dimgrey"))) +
  annotation_scale(bar_cols = c("dimgrey", "white"), # scale bar for both the main plot and inset
                   height = unit(0.15, "cm")) + 
  geom_shadowtext(data = group_ggh_pd,
                  aes(x = X, y = Y, label = REGION_name),
                  size = 2.5,
                  nudge_y = 3000,
                  nudge_x = 2000) +
  theme_minimal() +
  theme(legend.position = "none",
        axis.title = element_blank())

ggsave("images/TTS16-survey-area.png")
```
```{r, TTS-16-survey-area, echo=FALSE, fig.cap="\\label{fig:TTS-16-survey-area}The TTS 2016 study area within the Greater Golden Horseshoe in Ontario, Canada.", out.width="80%", fig.align='center'}
knitr::include_graphics("images/TTS16-survey-area.png")
```

```{r creating-desc-stats-table}
#forming a complete descriptive statistic table

Statistics <- data.frame("Statistics" = c("Min.", "1st Qu.", "Median", "Mean", "3rd Qu.", "Max.", "NA's"))

Trips <- data.frame("OD_Trips" = c(summary(od_ft_tt$trips)[[1]] %>% round(), 
                                   summary(od_ft_tt$trips)[[2]] %>% round(),  
                                   summary(od_ft_tt$trips)[[3]] %>% round(), 
                                   summary(od_ft_tt$trips)[[4]] %>% round(), 
                                   summary(od_ft_tt$trips)[[5]] %>% round(),
                                   summary(od_ft_tt$trips)[[6]]%>% round(),
                                   NA))

Travel_time <- data.frame("OD_Travel_time" = c(summary(od_ft_tt$travel_time)[[1]] %>% round(), 
                                               summary(od_ft_tt$travel_time)[[2]] %>% round(),  
                                               summary(od_ft_tt$travel_time)[[3]] %>% round(), 
                                               summary(od_ft_tt$travel_time)[[4]] %>% round(), 
                                               summary(od_ft_tt$travel_time)[[5]] %>% round(), 
                                               summary(od_ft_tt$travel_time)[[6]] %>% round(),  
                                               3507))

TAZ_Area <- data.frame("TAZ_Area" = c(summary(ggh_taz$AREA)[[1]] %>% round(), 
                                      summary(ggh_taz$AREA)[[2]] %>% round(), 
                                      summary(ggh_taz$AREA)[[3]] %>% round(), 
                                      summary(ggh_taz$AREA)[[4]] %>% round(), 
                                      summary(ggh_taz$AREA)[[5]] %>% round(), 
                                      summary(ggh_taz$AREA)[[6]] %>% round(), 
                                      NA))

@@

desc_stats <- cbind(Statistics, Trips, Travel_time, TAZ_Area, Workers, Jobs)

#kable tabling 
desc_stats %>%
  kable(format = "latex",
        align="lrrrrrr",
        booktabs = T,
        col.names = c(" ", "(#)", "(min)", "(km^2)", "(#)", "(#)"),
        caption = "\\label{tab:TTS-16-desc-stats}Descriptive statistics of the trips, workers, and jobs for the traffic analysis zones (TAZ) from the TTS 2016 dataset along with estimated car origin-destination travel times.") %>%
  add_header_above(c(" ", "Trips", "Car Travel Time", "Area", "Workers", "Jobs"), align = "r")%>%
  kable_styling(full_width = "T", 
                latex_options = c("scale_down"),
                position = "center")
```

@@

The origin-destination information consists of a cross-tabulation of people who are employed full-time by place of residence (origin) and places of employment in the GGH (destination) using the GTA06 zoning system. The number of workers and jobs is not equal; the boundaries of the survey are permeable, so workers who reside within the boundaries but travel outside of the boundaries are counted as workers within an origin TAZ, while jobs in TAZ that are filled by workers who reside outside the GGH boundaries are _unknown_ since they were not surveyed. This mismatch results in the total number of workers being `r round(sum(ggh_taz$workers)/sum(ggh_taz$jobs),2)` times larger than the number of jobs (i.e., `r sum(ggh_taz$workers)%>% prettyNum(big.mark = ",")` workers to `r sum(ggh_taz$jobs)%>% prettyNum(big.mark = ",")` jobs). While the 2016 TTS survey boundaries are drawn to minimize the difference between opportunities at the destination and supply at the origins they are still not equal <!-- cite? is this true ? -->. That said, these data offer a perspective of the potential of home-based trips to places of employment and as such, the number of trips taken are equal to the number of workers in the GGH.

```{r tts-workers-jobs-plot, fig.cap="\\label{fig:tts-workers-jobs-plot}Number of workers (top) and jobs (bottom) in each TAZ in the GGH area as specified in the 2016 TTS.", fig.width=8, fig.height=11}

tts_workers <- tm_shape(ggh_taz) +
  tm_fill(col = "workers", title = "Full-time employed people", palette = "-Spectral", style = "jenks") +
  tm_compass(type = "arrow", position = c("left", "top"), size = 2) +
  tm_scale_bar(breaks = c(0, 30, 60), text.size = 0.5, position = c("left", "bottom")) +
  tm_layout(legend.position = c("right", "bottom"))

tts_jobs <- tm_shape(ggh_taz) +
  tm_fill(col = "jobs", title = "Full-time jobs", palette = "-Spectral", style = "jenks") +
  tm_compass(type = "arrow", position = c("left", "top"), size = 2) +
  tm_scale_bar(breaks = c(0, 30, 60), text.size = 0.5, position = c("left", "bottom")) +
  tm_layout(legend.position = c("right", "bottom"))

@@

```{r ECD-plot, fig.cap="\\label{fig:ECD-plot}The cumulative distribution of the number of jobs and workers per TAZ from the 2016 TTS data set. Light blue shaded ranges correspond to all cumulative proabilities where the number of workers per TAZ are larger than jobs per TAZ. "}

@@

ggplot(ecdf_data , aes(x=Number, col=Group)) + 
  # geom_segment(aes(x = 115, y = 0, xend = 115, yend = 0.34), col = "blue", linetype=2) +
  # geom_segment(aes(x = 3055, y = 0, xend = 3055, yend = 0.94), col = "blue", linetype=2) +
  geom_rect(data = rect1, aes(xmin=xmin, xmax=xmax, ymin=ymin, ymax=ymax), fill="lightblue", alpha=0.2, inherit.aes = FALSE) +
  stat_ecdf() + 
  scale_x_continuous(trans='sqrt', breaks=c(115, 3055, 10000, 20000, 30000, 40000)) + 
  scale_y_continuous(breaks=c(0, 0.25,0.34, .50, .75,0.94, 1.00)) +
  labs(x = "Opportunities per TAZ",
       y = "Cumulative probability") + 
      scale_color_manual("Opportunities", 
                         values = c("Jobs" = "Black",
                                    "Workers" = "Red"))+
  scale_linetype_manual("Opportunities",values=c(0,1)) +
  geom_point(aes(x = 115, y = 0.34), col = "blue")+
  geom_point(aes(x=3055, y=0.94), col = "blue") +
  theme_minimal() +
  theme(plot.title = element_text(hjust=0.5),
        legend.position = c(0.90,0.75),
        axis.line = element_line(colour = "black"),
        panel.border = element_blank()) 
```

Figure \ref{fig:tts-workers-jobs-plot} presents the number of workers and jobs per TAZ. It can be observed that the spatial distribution of jobs and workers is unequal, which is indicative of jobs-housing imbalance, which can impact accessibility in a region [@Levine1998rethinking]. Workers are concentrated in TAZ within the center of the GGH and along the south-east and northern boarder of the GGH. The center of the GGH corresponds to the Greater Toronto Area (GTA) which is the most densely populated area in southern Ontario [@statistics_canada_daily_2022]. The south east border of the GGH neighbours Lake Ontario and is delineated by the urban built boundary of the Ontario Growth Plan being home to the highest density of working population in the GGH [@ontario_built_2019;@auditor_general_of_ontario_value_2021]. The northern GGH border corresponds to the Simcoe, Dufferin, Kawartha Lake, and Peterborough regions which are home to lower density of worker population density population [@auditor_general_of_ontario_value_2021]. Conversely, the spread of jobs in the GGH is lower than the number of workers indicating population is more spatially dispersed than jobs.

It can also be seen that from the bottom plot in Figure \ref{fig:tts-workers-jobs-plot} that high to medium-low concentrations of jobs are often present in the same areas as workers but only when the scale is transformed. In other words, though there is a higher number of TAZ with no workers than zones with no jobs (i.e., `r ggh_taz %>% st_drop_geometry() %>% count(workers) %>% filter(workers == 0) %>% pull('n')` TAZ with no workers : `r ggh_taz %>% st_drop_geometry() %>% count(jobs) %>% filter(jobs == 0) %>% pull('n')` TAZ with no jobs) and the mean of workers per TAZ is higher than the mean of jobs (i.e., `r round(mean(ggh_taz$workers, na.rm=T),0)` workers : `r round(mean(ggh_taz$jobs, na.rm=T),0)` jobs) the number TAZ with an extreme number of jobs at the highest and lowest percentiles is significantly higher than the number of workers; see the following cumulative probability distribution in Figure \ref{fig:ECD-plot} in which the 94th to 100th percentile and the 0th to 34th percentile of jobs in TAZ is higher than the number of workers in TAZ. This means that between these ranges, TAZ have a higher number of workers than they do jobs, echoing the more even spatial distribution of workers observed in Figure \ref{fig:tts-workers-jobs-plot}.

@@

{TTS2016R} also includes travel time data for each home-to-work trip as displayed in Figure \ref{fig:plot-tt-ttpertrip}. This travel time corresponds to a car commute calculated using the R package {r5r} and is interpreted as the travel time for a work commute for full-time employed people in the GGH. The travel times were calculated assuming the following input parameters: a maximum travel time less than or equal to 180 mins (3 hrs) and a street network retrieved from OpenStreetMaps for the GGH area. The 3 hr threshold was selected as it captures 99% of the trip taken (see the travel times summarized in Table \ref{tab:TTS-16-desc-stats}).

It is important to note that all travel times within this data set are calculated assuming car travel and one departure time for all origins. These assumptions are not completely realistic since we know a small proportion of trips are taken by non-car modes and travel time departures varies, however, it is not possible from the data retrieval system to obtain higher order tabulations that would allow us to split the data further. Thus we do not know which trips are made with non-car modes nor exact departure times in these tables. Though modal split and travel times can be estimated through other methods [e.g., @allen_suburbanization_2021; @higgins2021changes], for simplicity, we assume that all trips are taken by one-time departure car trip.

@@

```{r plot-tt-ttpertrip, fig.cap="\\label{fig:plot-tt-ttpertrip}Estimated total worker travel time (top) and average worker travel time (bottom) for each TAZ in the GGH.", fig.width=8, fig.height=11, message=FALSE}

@@

As can be observed in Figure \ref{fig:plot-tt-ttpertrip}, the total travel time (min) resembles the spatial trend distribution in the number of employed people in the previous plot (Figure \ref{fig:tts-workers-jobs-plot}). However, when the average travel time per trip in each TAZ is presented, the spatial distribution is distinct from all other plots presented so far. We can see that in areas around the south-eastern border that make up the Greater Toronto and Hamilton Area (GTHA) (e.g., Hamilton, Halton, Peel, Toronto, York, Durham) and Niagara and Waterloo, the average travel times are moderately low. Further from these areas, travel times are higher. Interestingly, even in eastern areas (e.g., Peterborough) with high employment and high job concentration, average travel time is higher than within the GTHA.

@@

Impedance functions are useful to understand mobility behavior and are part, for instance, of gravity models of spatial interaction [@wilson1971; @haynes_gravity_1985] and accessibility analysis in many applications [@hansen_how_1959;@levinson_accessibility_1998;@talen_assessing_1998;@reggiani_accessibility_2011;@paez_jobs_2013;@kuai_examining_2017; @barboza_balancing_2021]. An origin destination matrix and a cost matrix (i.e., with travel times) can be used in combination to estimate impedance functions. An impedance function $f(\cdot)$ depends on the cost of travel between locations $i$ and $j$ $c_{ij}$; it usually is a monotonically decreasing function, although sometimes the function can increase to reflect patterns of separation between activities; for instance, separation of land uses means that very short commuting trips are relatively rare. There can be sometimes also fluctuations due to hierarchical patterns, where travelers bypass opportunities in favor of more distant destinations that offer economies of agglomeration.

>>>>

---
title: "TTS2016R: A data set to study population and employment patterns from the 2016 Transportation Tomorrow Survey (TTS) in the Greater Toronto and Hamilton Area in Canada"
author:
  - name: "Anastasia Soukhov"
    email: "soukhoa@mcmaster.ca"
    affiliation: SEES
  - name: "Antonio Paez"
    email: "paezha@mcmaster.ca"
    affiliation: SEES
    footnote: 1
address:
  - code: SEES
    address: "School of Earth, Environment and Society, McMaster University, Hamilton, ON, L8S 4K1, Canada"
footnote:
  - code: 1
    text: "Corresponding Author"
abstract: 
  "This paper describes and visualises the data contained within the {TTS2016R} data-package created in `R`, the statistical computing and graphics language. In addition to a synthetic example, {TTS2016R} contains home-to-work commute information for the Greater Golden Horseshoe area in Canada retrieved from the 2016 Transportation Tomorrow Survey (TTS). Included are all Traffic Analysis Zones (TAZ), the number of people who are employed full-time per TAZ, the number of jobs per TAZ, origin-destination trips, and calculated car travel time from TAZ origin-destination centroid pairs. To illustrate how this information can be analysed to understand patterns in commuting, we estimate a distance-decay curve (i.e., impedance function) for the region. {TTS2016R} can be freely downloaded and explored at: https://github.com/soukhova/TTS2016R where the documentation and code involved in data creation, manipulation, and the final data products are detailed."

@@

```{r load-packages, include=FALSE, cache=FALSE}
library(TTS2016R)
library(dplyr)
library(fitdistrplus)
library(ggplot2)
library(kableExtra)
library(patchwork)
library(sf)
library(scales)
library(stats)
library(ggspatial)
library(shadowtext)
library(tmap)
library(ggpmisc)
# library(ggpmisc)
# library(ggrepel)
# library(cowplot)
# library(spdep)
# library(RColorBrewer)
# library(extrafont)
# font_import()
# loadfonts(device = "win")

@@

This manuscript presents the open data product [{TTS2016R}](https://github.com/soukhova/TTS2016R). Open data products are the result of turning source data (open or otherwise) into accessible information that adds value to the original inputs [see @Arribas2021open]. The product presented in this paper is an `R` data package which currently consists of three objects which are sourced from the 2016 Transportation Tomorrow Survey (TTS) or are curated to facilitate the use and analysis of TTS data. This package includes person-to-jobs origin-destinations, traffic analysis zone (TAZ) boundaries, and planning/municipality boundaries for the Greater Golden Horse area (GGH) located in southern Ontario, Canada [@data_management_group_tts_2018]. In addition, the package includes TAZ centroid-to-centroid travel times by car computed using package {r5r} [@Pereira2021r5r]. The aim of this paper is to walk readers through the empirical home-based work commute data set, illustrate the calculation of an impedance function that can be used to calculate accessibility to employment, and invite its use in other applications.

<!-- Data from the TTS are in principle available to the public but are not fully open, since permission to access the data retrieval system is required. In addition, the raw data can be technically demanding, cumbersome to work with, and requires multiple software to process. By pre-processing the data in a `R` environment, {TTS2016R} offers a slice of the TTS data useful to understand patterns of commuting to work in the region. It also provides open infrastructure for additional TTS or complimentary data sets to be amended by the authors or wider open-source community in the future. -->

@@

{TTS2016R} includes counts of fully-employed population by place of residence (origin) and counts of full-time jobs by place of work (destination) aggregated by TAZ (n=`r round(length(TTS2016R::ggh_taz$GTA06), 3) %>% prettyNum(big.mark = ",")` within the survey boundaries). TAZ typically are defined based on land-use and population demographics in order to estimate the number of trips produced and attracted to each zone [@meyer_urban_2001]. As such, each TAZ is uniquely identified using the GTA06 Zoning System which can be used to join to the origin-destination table (i.e., trips taken).

The number of jobs (`r round(sum(TTS2016R::ggh_taz$jobs), 3) %>% prettyNum(big.mark = ",")`) and workers (`r round(sum(TTS2016R::ggh_taz$workers), 3) %>% prettyNum(big.mark = ",")`) in this package are organized in the form of an origin-destination table which is indicative of home-to-work commute patterns (there are `r round(sum(TTS2016R::od_ft_tt$trips), 3) %>% prettyNum(big.mark = ",")` potential interactions). These data were retrieved from the TTS Data Retrieval System on October 28, 2021 and reflect the potential interaction of full-time employed people and jobs within the GGH survey boundaries shown in Figure \ref{fig:TTS-16-survey-area} as defined by the 2016 TTS methodology [@data_management_group_tts_2018].

Also included in {TTS2016R} are travel times and cost of travel from origin to destination by car; travel times are calculated using the `R` package {r5r}. These travel times are useful to estimate the cost of travel and to calculate impedance functions, among other possible applications. For simplicity, all interactions within {TTS2016R} are assumed to be taken by car, and the travel time is calculated from an origin TAZ centroid to a destination TAZ centroid. The centroid is snapped to the nearest street line by `r5r` and the travel time is calculated for all trips assuming a car travel mode. Additionally, only travel times less than or equal to 180 mins (3 hrs) are calculated; this threshold represents 99% of trip's travel times which are summarized in Figure \ref{fig:TTS-16-survey-area}.

@@

group_ggh_pd[group_ggh_pd$REGION_name=="Peterborough County", "Y"] <- 4921000
```
```{r, echo=FALSE, message=FALSE, warning=FALSE, include = FALSE}
#plot 
TTS_area_plot <- ggplot() +
  geom_sf(data = TTS2016R::ggh_pd, color = "darkgray",
          aes(fill = REGION_name)) +
  scale_fill_manual(values = viridis::viridis(20)) +
  geom_sf(data = group_ggh_pd_poly, 
          color = "black", fill = NA, size = 0.7) +
  annotation_north_arrow(location = "tl", # north arrow for both the main plot
                         height = unit(0.8, "cm"), 
                         width = unit(0.8, "cm"),
                         style = north_arrow_orienteering(line_width = 0.25,
                                                          line_col = "dimgrey", 
                                                          fill = c("white","dimgrey"))) +
  annotation_scale(bar_cols = c("dimgrey", "white"), # scale bar for both the main plot
                   height = unit(0.15, "cm")) + 
  geom_shadowtext(data = group_ggh_pd,
                  aes(x = X, y = Y, label = REGION_name),
                  size = 3,
                  nudge_y = 3000,
                  nudge_x = 2000) +
  theme_void() +
  theme(legend.position = "none",
        axis.title = element_blank())

# ggsave("images/TTS16-survey-area.png")
```
```{r creating-desc-stats-table}
#forming a complete descriptive statistic table

Statistics <- data.frame("Statistic" = c("Min.", "1st Qu.", "Median", "Mean", "3rd Qu.", "Max.", "NA's"))

Trips <- data.frame("Trips" = c(summary(od_ft_tt$trips)[[1]] %>% round(), 
                                   summary(od_ft_tt$trips)[[2]] %>% round(),  
                                   summary(od_ft_tt$trips)[[3]] %>% round(), 
                                   summary(od_ft_tt$trips)[[4]] %>% round(), 
                                   summary(od_ft_tt$trips)[[5]] %>% round(),
                                   summary(od_ft_tt$trips)[[6]]%>% round(),
                                   NA))

Travel_time <- data.frame("TT" = c(summary(od_ft_tt$travel_time)[[1]] %>% round(), 
                                               summary(od_ft_tt$travel_time)[[2]] %>% round(),  
                                               summary(od_ft_tt$travel_time)[[3]] %>% round(), 
                                               summary(od_ft_tt$travel_time)[[4]] %>% round(), 
                                               summary(od_ft_tt$travel_time)[[5]] %>% round(), 
                                               summary(od_ft_tt$travel_time)[[6]] %>% round(),  
                                               3507))

# TAZ_Area <- data.frame("TAZ_Area" = c(summary(ggh_taz$AREA)[[1]] %>% round(1), 
#                                       summary(ggh_taz$AREA)[[2]] %>% round(1), 
#                                       summary(ggh_taz$AREA)[[3]] %>% round(1), 
#                                       summary(ggh_taz$AREA)[[4]] %>% round(1), 
#                                       summary(ggh_taz$AREA)[[5]] %>% round(1), 
#                                       summary(ggh_taz$AREA)[[6]] %>% round(1), 
#                                       NA))

@@

desc_stats <- cbind(Statistics, Trips, Travel_time, Workers, Jobs)

# #kable tabling 
# desc_stats %>%
#   kable(format = "latex",
#         align="lrrrrrr",
#         booktabs = T,
#         col.names = c(" ", "(#)", "(min)", "(km^2)", "(#)", "(#)"),
#         caption = "\\label{tab:TTS-16-desc-stats}Descriptive statistics of the trips, calculated origin-destination car travel time, TAZ area, workers per TAZ, and jobs per TAZ.") %>%
#   add_header_above(c(" ", "Trips", "Travel Time", "Area", "Workers", "Jobs"), align = "r")%>%
#   kable_styling(full_width = "T", 
#                 latex_options = c("scale_down"),
#                 position = "center")
```

```{r, TTS-16-survey-area, echo=FALSE, fig.cap="\\label{fig:TTS-16-survey-area}The TTS 2016 study area within the GGH in Ontario, Canada along with the descriptive statistics of the trips, calculated origin-destination car travel time (TT), workers per TAZ, and jobs per TAZ. Contains 20 regions (black boundaries) and sub-regions (dark gray boundaries).", out.width="100%", fig.align='center'}
# knitr::include_graphics("images/TTS16-survey-area.png")

TTS_area_plot + annotate(geom = "table",
            x = 921500, y = 4890000, 
           label = list(desc_stats))
```

@@

The origin-destination table (i.e., trips) consists of a cross-tabulation of people who are employed full-time by place of GGH residence (origin) and places of GGH employment (destination) using the GTA06 TAZ zoning system. It is important to note that the total number of workers and jobs is the TTS 2016 region are not equal but the number of trips taken are equal to the number of workers. Since the outer boundaries of the TTS are permeable, workers who reside within the boundaries but travel outside of the boundaries are counted as workers within an origin TAZ, while jobs in TAZ that are filled by workers who reside outside the GGH boundaries are _unknown_ since they were not surveyed. This mismatch results in the total number of workers being `r round(sum(ggh_taz$workers)/sum(ggh_taz$jobs),2)` times larger than the number of jobs (i.e., `r sum(ggh_taz$workers)%>% prettyNum(big.mark = ",")` workers to `r sum(ggh_taz$jobs)%>% prettyNum(big.mark = ",")` jobs). As such, the origin-destination table contained in {TTS2016R} offers a perspective on all workers in the GGH and their home-based trips to places of GGH employment.

```{r tts-workers-jobs-plot, fig.cap="\\label{fig:tts-workers-jobs-plot}Number of workers (left) and jobs (right) in each TAZ in the GGH area as specified in the 2016 TTS.", fig.width=8, fig.height=5}

tts_workers <- tm_shape(ggh_taz) +
  tm_fill(col = "workers", title = "Full-time \nemployed people", palette = "-Spectral", style = "jenks") +
  tm_compass(type = "arrow", position = c("left", "top"), size = 2) +
  tm_scale_bar(breaks = c(0, 30, 60), text.size = 0.5, position = c("left", "bottom")) +
  tm_layout(legend.position = c("right", "bottom"))

tts_jobs <- tm_shape(ggh_taz) +
  tm_fill(col = "jobs", title = "Full-time \njobs", palette = "-Spectral", style = "jenks") +
  tm_compass(type = "arrow", position = c("left", "top"), size = 2) +
  tm_scale_bar(breaks = c(0, 30, 60), text.size = 0.5, position = c("left", "bottom")) +
  tm_layout(legend.position = c("right", "bottom"))

@@

<!--
```{r ECD-plot, fig.cap="\\label{fig:ECD-plot}The cumulative distribution of the number of jobs and workers per TAZ from the 2016 TTS data set. Light blue shaded ranges correspond to all cumulative proabilities where the number of workers per TAZ are larger than jobs per TAZ.", fig.width=8, fig.height=4}

@@

ggplot(ecdf_data , aes(x=Number, col=Group)) + 
  # geom_segment(aes(x = 115, y = 0, xend = 115, yend = 0.34), col = "blue", linetype=2) +
  # geom_segment(aes(x = 3055, y = 0, xend = 3055, yend = 0.94), col = "blue", linetype=2) +
  geom_rect(data = rect1, aes(xmin=xmin, xmax=xmax, ymin=ymin, ymax=ymax), fill="lightblue", alpha=0.2, inherit.aes = FALSE) +
  stat_ecdf() + 
  scale_x_continuous(trans='sqrt', breaks=c(115, 3055, 10000, 20000, 30000, 40000)) + 
  scale_y_continuous(breaks=c(0, 0.25,0.34, .50, .75,0.94, 1.00)) +
  labs(x = "Opportunities per TAZ",
       y = "Cumulative probability") + 
      scale_color_manual("Opportunities", 
                         values = c("Jobs" = "Black",
                                    "Workers" = "Red"))+
  scale_linetype_manual("Opportunities",values=c(0,1)) +
  geom_point(aes(x = 115, y = 0.34), col = "blue")+
  geom_point(aes(x=3055, y=0.94), col = "blue") +
  theme_minimal() +
  theme(plot.title = element_text(hjust=0.5),
        legend.position = c(0.90,0.75),
        axis.line = element_line(colour = "black"),
        panel.border = element_blank()) 
```
-->
Figure \ref{fig:tts-workers-jobs-plot} presents the number of workers and jobs per TAZ. It can be observed that the spatial distribution of jobs and workers is unequal, which is indicative of a jobs-housing imbalance that can impact accessibility in a region [@Levine1998rethinking]. It can also be seen that there is a higher number of TAZ with no workers than zones with no jobs (i.e., `r ggh_taz %>% st_drop_geometry() %>% count(workers) %>% filter(workers == 0) %>% pull('n')` TAZ with no workers : `r ggh_taz %>% st_drop_geometry() %>% count(jobs) %>% filter(jobs == 0) %>% pull('n')` TAZ with no jobs) and the mean of workers per TAZ is higher than the mean of jobs (i.e., `r round(mean(ggh_taz$workers, na.rm=T),0)` workers : `r round(mean(ggh_taz$jobs, na.rm=T),0)` jobs) the number TAZ with an extreme number of jobs at the highest and lowest percentiles is significantly higher than the number of workers. <!-- see the following cumulative probability distribution in Figure \ref{fig:ECD-plot}. -->

<!-- Workers are concentrated in TAZ within the center of the GGH and along the south-east and northern boarder of the GGH. The center of the GGH corresponds to the Greater Toronto Area (GTA) which is the most densely populated area in southern Ontario [@statistics_canada_daily_2022]. The south east border of the GGH neighbours Lake Ontario and is delineated by the urban built boundary of the Ontario Growth Plan being home to the highest density of working population in the GGH [@ontario_built_2019;@auditor_general_of_ontario_value_2021]. The northern GGH border corresponds to the Simcoe, Dufferin, Kawartha Lake, and Peterborough regions which are home to lower density of worker population density population [@auditor_general_of_ontario_value_2021]. Conversely, the spread of jobs in the GGH is lower than the number of workers indicating population is more spatially dispersed than jobs. -->

<!--- in which the 94th to 100th percentile and the 0th to 34th percentile of jobs in TAZ is higher than the number of workers in TAZ. This means that between these ranges, TAZ have a higher number of workers than they do jobs, echoing the more even spatial distribution of workers observed in Figure \ref{fig:tts-workers-jobs-plot}. -->

@@

As mentioned, {TTS2016R} also includes travel time data for each home-to-work trip as displayed in Figure \ref{fig:plot-tt-ttpertrip}. This travel time corresponds to a car commute calculated using the R package {r5r} (see descriptive statistics in Table \ref{tab:TTS-16-desc-stats}). It is important to note that travel times within this data set are calculated assuming only car travel and one departure time for all origins. These assumptions are not completely realistic since we know a small proportion of trips are taken by non-car modes and travel time departures varies. However, it is not possible from the data retrieval system to obtain higher order tabulations so <!--  we do not know which trips are made with non-car modes nor exact departure times in these tables. Though modal split and travel times can be estimated through other methods [e.g., @allen_suburbanization_2021; @higgins2021changes], for simplicity,-->  we carry on with the assume that all trips are taken by one-time departure car trip.

@@

```{r plot-tt-ttpertrip, fig.cap="\\label{fig:plot-tt-ttpertrip}Estimated total worker travel time (left) and average worker travel time (right) for each TAZ in the GGH.", fig.width=8, fig.height=5, message=FALSE}

@@

As can be observed in Figure \ref{fig:plot-tt-ttpertrip}, the total travel time resembles the spatial trend distribution in the number of employed people in the previous plot (Figure \ref{fig:tts-workers-jobs-plot}) and the spatial distribution of the average travel time is distinct from other plots presented so far. We can see that in areas around the south-eastern border that make up the Greater Toronto and Hamilton Area (GTHA)<!-- (e.g., Hamilton, Halton, Peel, Toronto, York, Durham) -->, Niagara and Waterloo, the average travel times are moderately low. Further from these areas, travel times are higher. Interestingly, even in eastern areas (e.g., Peterborough) with high employment and high job concentration, average travel time is higher than within the GTHA.

@@

Impedance functions are useful to understand mobility behaviour and are used to estimate gravity models of spatial interaction [@wilson1971; @haynes_gravity_1985] and applied in accessibility analysis [@hansen_how_1959;@paez_jobs_2013;@barboza_balancing_2021]. An impedance function $f(\cdot)$ depends on the cost of travel $c_{ij}$ between locations $i$ and $j$ (all which is supplied in the travel time and origin-destination table within {TTS2016R}). <!-- The impedance function is usually a monotonically decreasing function since the separation of land uses means that very short commuting trips are relatively rare. However, sometimes the function can increase to reflect patterns of separation between activities. This could be needed to reflect fluctuations in travel cost due to hierarchical patterns, where travelers bypass opportunities in favor of more distant destinations that offer economies of agglomeration. -->
